### 2024-06-05 rebuttals/rebuttals.tex

\documentclass[12pt,a4paper]{article}
\usepackage{longtable}
\usepackage[LGR,T1]{fontenc}
\usepackage{lscape}
%\usepackage{pdfsync}
\usepackage{multirow}
\usepackage{amsmath,bm} 
\usepackage{amsfonts}
\usepackage{amsthm} % Extended theorem environments
%\usepackage{amssymb} % Math symbols %%redundant with stix package
%\usepackage{esint} % Intégrales multiples
%\usepackage{esvect} % Vecteurs
\usepackage{mathtools}
\usepackage{pifont}

\usepackage{fancyhdr}
\usepackage{graphicx}

\graphicspath{{../frontend/img/}{../chap_intro_ccl/img/}{../chap_case_study/img/}{../chap_ES_PCE/img/}{../chap_electro_uq/img/}{../chap_methodo/img/}{../chap_myopic/img/}{../chap_RL/img/}{../chap_atom_mol/img/}{../chap_RobPol/img/}{../appendices/img/}} % Figures folder different for each input
%\graphicspath{{img/}} %
\DeclareGraphicsExtensions{.eps,.pdf,.png,.jpg}

\usepackage{lastpage}
\usepackage{afterpage}
\usepackage{lettrine}
\usepackage{color,soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{colortbl}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{titlesec}
\def\eg{e.g.,\ }
\def\ie{i.e.,\ }

%Palatino font
%\usepackage{pxfonts}
%\usepackage{libertine}
\usepackage[scaled=0.88]{beraserif}
\usepackage[scaled=0.85]{berasans}
\usepackage[scaled=0.84]{beramono}
\usepackage{mathpazo}
%\linespread{1.05}
\usepackage[T1,small,euler-digits]{eulervm}

\usepackage[nomessages]{fp}

\definecolor{bleuUCLclair}{rgb}{.09, 0.569, 1}
\definecolor{bleuUCLfonce}{rgb}{ .13, .52, .86}
\definecolor{redBurn}{rgb}{.91, 0.29, 0.08}

\usepackage{tocloft}
\renewcommand{\cftsecpresnum}{Reviewer \#}
\renewcommand{\cftsecnumwidth}{6em}
\renewcommand{\cftsubsecpresnum}{Comment \#}
\renewcommand{\cftsubsecnumwidth}{6em}

\usepackage[colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black]{hyperref}
\usepackage[square,numbers,sort&compress]{natbib}

\bibliographystyle{biblio/elsarticle-num-names}  %Ordered by appearance in the text, with DOI and URL
%\usepackage[backend=biber, natbib=true, style=numeric-comp, citestyle=numeric-comp, sorting=none, giveninits=true, maxcitenames=1]{biblatex}

\usepackage[]{tocbibind}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
%    linkcolor=black,
    bookmarks=true,
    pdfpagemode=FullScreen,
}
\setcounter{tocdepth}{1}

\addtolength{\topmargin}{-1.5cm}
\addtolength{\textheight}{1.5cm}
\addtolength{\textwidth}{2cm}
\addtolength{\footskip}{2cm}
\setlength{\evensidemargin}{-0.5cm}
\setlength{\oddsidemargin}{-0.5cm}
\setlength{\arrayrulewidth}{0.25pt}

\renewcommand{\baselinestretch}{1.1} % Interligne

\newenvironment{maliste}%
{ \begin{list}%
	{\textcolor{bleuUCLfonce}{$\bullet$}\hspace{0.5cm}}%
	{\setlength{\labelwidth}{50pt}%
	 \setlength{\leftmargin}{25pt}%
	 \setlength{\itemsep}{30pt}}}%
{ \end{list} }

%\renewcommand{\headrulewidth}{0.0pt}
%\newcommand{\clearemptydoublepage}{%
%	\newpage{\pagestyle{empty}\cleardoublepage}}

%section like title in longtable
\newcommand{\seclong}[1]{\multicolumn{2}{@{}l}{{\Large\sffamily #1}} 
\vspace{0.5cm}
\\}

%enumerate on two columns
\newcounter{listlong}
\newcommand{\newlistlong}{\setcounter{listlong}{1}}
\newcommand{\iteml}[1]{%
\hspace{4.5cm}\textcolor{redBurn}{\arabic{listlong}}\stepcounter{listlong}%
&%
#1%

\\%
}

%left in column
\newcommand{\lcol}[1]{%
\begin{minipage}[t]{.35\textwidth}%

#1%

\end{minipage}%
}

\title{\vspace{-1cm}
\begin{flushleft} {\sffamily Xavier Rixhon's PhD thesis - Answers to jury members}\end{flushleft}}
\date{\vspace{-1.7cm}\begin{flushleft}\sffamily Exploration of uncertainty-aware energy transition pathways - Reinforcement learning and principal component analysis-based methods\end{flushleft}}
%
%Xavier Rixhon, Gauthier Limpens, Diederik Coppitters, Hervé Jeanmart and Francesco Contino\end{flushleft}}

\pagestyle{fancy} 
\fancyhf{}
\fancyfoot[R]{\sffamily\thepage\ / \pageref{LastPage}}
  \fancyfoot[L]{ }

\fancypagestyle{plain}{%
  \fancyhf{}%
  \fancyfoot[R]{\sffamily\thepage\ / \pageref{LastPage}}
  \fancyfoot[L]{ }
}

\renewcommand{\headrulewidth}{0.0pt}

\newcommand{\hlc}[2][yellow]{ {\sethlcolor{#1} \hl{#2}} }

\titleformat{\section}
  {\bfseries\scshape}{}{1em}{}

\titleformat{\subsection}
  {\normalfont\scshape}{}{1em}{}

\usepackage[framemethod=default]{mdframed}
%\mdfsetup{skipabove=\topskip,skipbelow=\topskip}

\global\mdfdefinestyle{comment}{%
     linecolor=red,linewidth=0.1cm,%
     leftmargin=-0.5cm,rightmargin=-0.5cm, innerleftmargin=0.4cm,innerrightmargin=0.4cm,
     topline=false,bottomline=false
}

\global\mdfdefinestyle{manuscript}{%
     linecolor=gray!20,linewidth=0.05cm,backgroundcolor=gray!20,%
     leftmargin=-0.5cm,rightmargin=-0.5cm, innerleftmargin=0.4cm,innerrightmargin=0.4cm
}

  \renewcommand{\subsectionautorefname}{Comment}
\begin{document}
\maketitle
%\emph{Start with some thank you note. For example:}

I would like to thank the jury members for the comments, which significantly helped improving the manuscript and substantiating the novelty of my work. Based on the notes taken during the private defense, I hereby transcribed, as accurately as possible, the jury's comments.

I believe I have addressed all the issues raised in the following answers. Some of them required adaptations of the text. These adaptations are either directly in the thesis manuscript or left for further developments in subsequent papers. For each comment, I have first highlighted the issue, then provided an answer, and finally described how the manuscript was adjusted, if needed.\\

When a comment explicitly came from specific members of the jury, they are listed at the beginning of the comment with the following color code: {\color{orange} \textbf{Stefano} Moret}, {\color{teal} \textbf{Stefan} Pfenninger}, {\color{purple} \textbf{Sylvain} Quoilin} and {\color{violet} \textbf{Christophe} De Vleeschouwer}. Here is how an answer to a comment is structured:

\begin{mdframed}[style=comment] % Comment from the reviewer
Text of the comment.
\end{mdframed}

\noindent The answer provided to the comment and, based on this, where the potential modification brought to the manuscript is located in the thesis manuscript using {\color{blue} blue font}.

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
New version of the text in the manuscript.
\end{mdframed}

\clearpage

%
% Reviewer 1
%
\section{Introduction}
\label{Introduction}
%Thank you for the reviews and comments to the manuscript we submitted to the Frontiers Journal. In the following, we carefully addressed each point raised by Reviewer 1, and revised the main manuscript and the supporting information accordingly. The following replies include as well all the changes we made to our manuscript based on your comments.

\subsection{Literature review}
\label{Intro_lit_review}

\section{Reinforcement Learning}
\label{RL}
%Thank you for the reviews and comments to the manuscript we submitted to the Frontiers Journal. In the following, we carefully addressed each point raised by Reviewer 1, and revised the main manuscript and the supporting information accordingly. The following replies include as well all the changes we made to our manuscript based on your comments.

\subsection{Binding}
\label{RL_binding}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} - The concept of binding constraint was not clear.
\end{mdframed}

\noindent
Besides the information given in Section 4.2.3 of the manuscript (and reminded here below), I do not see further explanations that could clarify the concept of binding constraint in a Linear Programming (LP) problem. Consequently, regarding this comment, there has not been further modification brought to the manuscript.

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
To identify the actions that have an actual impact on the environment, we can check if they are binding or not. In a LP problem, constraints represent hyperplanes in the domain of variables. In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}). When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility. The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function. Besides being within the domain of feasibility, it is proven that this optimal solution, when unique\footnote{There are cases where the objective function has the same optimal value along an entire edge. In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}. The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal. In other words, binding constraints, when tightened, aggravate the objective value function. If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.
\end{mdframed}

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
\caption{Binding versus non-binding constraints. In LP where the feasibility domain is non-empty and bounded, the constraints defined a convex feasibility domain in the space of variables (here, x$_1$ and x$_2$). The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution. These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
\label{fig:Binding_constr}
\end{figure}

\section{Principal Component Analysis}
\label{PCA}

\subsection{Confusion with the word PC}
\label{Confusion_PC_wording}

\begin{mdframed}[style=comment] % Comment from the reviewer
{\color{violet} \textbf{Christophe}} When using the word ``component'', there seems to be a confusion and it is not always easy to understand if you refer to the vector or the coefficient related to one of the original variable.
\end{mdframed}

\noindent The confusion probably comes from the fact a Principal Component (PC) actually represents an eigenvector of the covariance matrix. This vector is, by definition, composed of \textbf{components}, each of them being a coefficient related to a specific original variable. Here below the parts where there could have been misleading confusion are listed and, potentially, modifications have been brought to the manuscript:

{\color{blue} End of second paragraph of section 1.4.1 - no modification}:

\begin{mdframed}[style=manuscript] % Modification brought to the manuscript
Moreover, this means that $\alpha_{ki}$, \ie the \textbf{component} of $\bm{\alpha}_{\mathbf{k}}$ related to the $i^{\text{th}}$ original variable, $x_i$,  gives its weight in the $k^{\text{th}}$ PC, \ie $z_k$. 
\end{mdframed}

\clearpage
\def\bibfont{\scriptsize}
\bibliography{../bib_thesis.bib}
\normalsize

\end{document}

### 2024-06-05 chap_RL/chap_RL.tex
Typos + Init Rebuttals

Uncertainties about the future along with a large variety of \gls{IAMs} yield to an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}. For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century. On the contrary, using his top-down model DICE, \citet{nordhaus2014question} states that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.

@@

To identify the actions that have an actual impact on the environment, we can check if they were binding or not. In a \gls{LP} problem, constraints represent hyperplanes in the domain of variables. In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}). When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility. The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function. Besides being within the domain of feasibility, it is proven that this optimal solution, when unique\footnote{There are cases where the objective function has the same optimal value along an entire edge. In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}. The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal. In other words, binding constraints, when tightened, aggravate the objective value function. If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.325\textwidth]{Opex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Capex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
\caption{Comparison of cumulative OPEX (left), CAPEX (right) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
\label{fig:Opex_Capex_Salvage_comp}
\end{figure}

>>>>

Uncertainties about the future along with a large variety of \gls{IAMs} yield an even larger variety of \gls{GHG} emissions reduction pathways \cite{nicolas2021robust}. For instance, several studies \cite{IPCC_CO2_budget,steffen2018trajectories} advocate for actions to take in the near future, especially to keep on track within the 1.5°C (if not, 2°C) increase of global temperature above pre-industrial levels by the end of the century. On the contrary, using his top-down model DICE, \citet{nordhaus2014question} states that immediate and drastic actions are not compulsory to meet the ambition of climate change mitigation.

@@

To identify the actions that have an actual impact on the environment, we can check if they are binding or not. In a \gls{LP} problem, constraints represent hyperplanes in the domain of variables. In a two-dimension space, these are straight lines (see Figure \ref{fig:Binding_constr}). When the problem is bounded and feasible, these lines are the edges of a convex polygon: the domain of feasibility. The optimal solution, $\textbf{x}^*$, is the combination of variables leading to the optimal value of the objective function. Besides being within the domain of feasibility, it is proven that this optimal solution, when unique\footnote{There are cases where the objective function has the same optimal value along an entire edge. In this case, there is an infinity of solutions and the problem is indeterminate.}, locates on a vertex of the domain \cite{bertsimas1997introduction}. The constraints intersecting at this vertex are considered as binding, actually limiting the objective function to be more optimal. In other words, binding constraints, when tightened, aggravate the objective value function. If these are inequality constraints, as represented in Figure \ref{fig:Binding_constr}, it means that their left and right sides of the equation are equal.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.325\textwidth]{Opex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Capex_2050_comp_core.pdf}
\includegraphics[width=0.325\textwidth]{Salvage_2050_comp_core.pdf}
\caption{Comparison of cumulative OPEX (left), CAPEX (center) and salvage value (right) in 2050 from the perfect foresight optimisation under uncertainties and the \gls{RL}-based myopic optimisation.}
\label{fig:Opex_Capex_Salvage_comp}
\end{figure}

### 2024-06-05 chap_RobPol/chap_RobPol.tex
Typos + Init Rebuttals

\newpage
\section{Robustness assessment of pathway roadmaps}
\label{sec:RobPol:Rob_Assessment}
Now that the performance metric is defined, we can assess the robustness of different roadmaps. These roadmaps are defined as the technological mix given by the deterministic optimisation of the perfect foresight pathway under certain conditions. The two first roadmaps are the \textbf{REF} and \textbf{SMR} cases (see Chapter \ref{chap:atom_mol}). All the uncertain parameters are considered at their nominal value in these two cases. The only difference consists in allowing \gls{SMR} from 2040 onward in the \textbf{SMR} case. On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''. Then, we assess the robustness of these three roadmaps via a ``three-step rolling horizon approach'' extended from the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.

@@

\begin{table}[htbp!]
\caption{Comparison of the robustness with the REF case where robustness is defined by the width of the 95\% confidence range of the additional installed capacities projected on the PC. Differences below 5\% are not shown ($\simeq$).}
\label{tab:projection}
\centering
\begin{tabular}{c l| c c}
\toprule
\multirow{2}{*}{\textbf{PC}} & \multirow{2}{*}{\textbf{Main contributors}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
& & SMR & ROB \\
\midrule
1 & \gls{PV} and electrification & $\simeq$ & +49\% \\
2 & \gls{PV} and dec.heating & $\simeq$ & +37\%\\
3 & Freight transport & $\simeq$ & +33\%\\
4 & Dec. oil boilers & $\simeq$ & $\simeq$ \\
5 &  Naphtha-crackers and \gls{MTO} & +34\% & -21\%\\
\bottomrule							
\end{tabular}
\end{table}

@@

@@

Then, we have tested three roadmaps resulting from different perfect foresight optimisations: REF, SMR and ROB. Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}. From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation of the additional capacities needed in myopic optimisations subject to uncertainties. This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation. However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario. Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).

>>>>

\newpage
\section{Robustness assessment of pathway roadmaps}
\label{sec:RobPol:Rob_Assessment}
Now that the performance metric is defined, we can assess the robustness of different roadmaps. These roadmaps are defined as the technological mix given by the deterministic optimisation of the perfect foresight pathway under certain conditions. The two first roadmaps are the \textbf{REF} and \textbf{SMR} cases (see Chapter \ref{chap:atom_mol}). All the uncertain parameters are considered at their nominal value in these two cases. The only difference consists in allowing \gls{SMR} from 2040 onward in the \textbf{SMR} case. On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''. Then, we assess the robustness of these three roadmaps via a ``three-step rolling horizon approach'' extended from the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of additional investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}. Finally, we assess the variation of the transition costs resulting from these myopic transitions under uncertainties.

@@

\begin{table}[htbp!]
\caption{Comparison of the robustness with the REF case where robustness is defined by the width of the 95\% confidence range of the additional installed capacities projected on the PC. Differences below 5\% are not shown ($\simeq$).}
\label{tab:projection}
\centering
\begin{tabular}{c l| c c}
\toprule
\multirow{2}{*}{\textbf{PC}} & \multirow{2}{*}{\textbf{Main contributors}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
& & SMR & ROB \\
\midrule
1 & \gls{PV} and electrification & $\simeq$ & +49\% \\
2 & \gls{PV} and dec. heating & $\simeq$ & +37\%\\
3 & Freight transport & $\simeq$ & +33\%\\
4 & Dec. oil boilers & $\simeq$ & $\simeq$ \\
5 &  Naphtha-crackers and \gls{MTO} & +34\% & -21\%\\
\bottomrule							
\end{tabular}
\end{table}

@@

\subsection{What about the costs over the transition?}
\label{subsec:RobPol:costs_comparison}
Besides the additional capacities needed through the myopic transitions, this section takes a step back to assess the variations of costs over the transition: the total transition cost, the cumulative \gls{OPEX} by 2050 and the cumulative \gls{CAPEX} by 2050 (see Figure \ref{fig:Total_Opex_Capex_REF_ROB}). We observe that the differences are marginal. This validates the initial rationale to go beyond these costs to assess the robustness of roadmaps.

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.325\textwidth]{Transition_cost_REF_ROB.pdf}
\includegraphics[width=0.325\textwidth]{OPEX_2050_REF_ROB.pdf}
\includegraphics[width=0.325\textwidth]{CAPEX_2050_REF_ROB.pdf}
\caption{Comparison of total transition cost (left), cumulative OPEX (center) and CAPEX (right) by 2050 from the REF and ROB myopic transitions. The differences are marginal. However, the ROB roadmap lead to cheaper and less varying total transition cost. Requiring more investments at earlier stages, this roadmap is less affected by the variation of the operational costs.}
\label{fig:Total_Opex_Capex_REF_ROB}
\end{figure}

Considering the median values, the ROB roadmap leads to a 1.6\% more expensive total transition. This comes from the earlier investments in \gls{PV} panels together with electrified solutions in the heating sector that need to be renewed before the end of the transition. The extra-\gls{CAPEX} also comes from more efficient (and more expensive) technologies such as \gls{CHP}. These decisions allow being less affected by the variation of the cumulative operational costs that are dominated ($\sim$\,70\%) by the cost of purchasing energy carriers.  This makes the investment-to-operation ratio slightly higher for the ROB case (46\%) than for the REF case (44\%).

We can assess the variation of these costs with the quartile coefficient of dispersion. This coefficient of dispersion is defined as $\left(Q_3-Q_1 \right)/\left(Q_3+Q_1 \right)$ where $Q_1=P_{25}$ and $Q_3=P_{75}$ are the first and third quartiles of the data, respectively. We observe that the ROB case has a smaller quartile coefficient of dispersion for the total transition cost and cumulative \gls{OPEX} but higher for the cumulative \gls{CAPEX} (see Table \ref{tab:quartile_coeff_dispersion}).The biggest decrease of these coefficients of dispersion between the REF and the ROB cases concerns the total transition cost. This means that the ROB case ensures to decrease more the uncertainty on this cost.

\begin{table}[htbp!]
\caption{Quartile coefficient of dispersion for the total transition cost, cumulative OPEX and CAPEX by 2050 from the REF and ROB myopic transitions. The ROB case leads to more certain total transition cost and cumulative \gls{OPEX} than the REF case.}
\label{tab:quartile_coeff_dispersion}
\centering
\begin{tabular}{l| c c}
\toprule
\textbf{Quartile coefficient of dispersion} &  REF  & ROB \\
\midrule
Total transition cost & 12.9\% & 12.2\%  \\
Cumulative \gls{OPEX} by 2050 & 21.3\% & 21.1\% \\
Cumulative \gls{CAPEX} by 2050 & 5.9\% & 6.0\%\\
\bottomrule							
\end{tabular}
\end{table}

All these conclusions are similar to the ones observed on the study of overcapacity of \citet{moret2020overcapacity}). Working on the snapshot model and fixing the installed capacities only in the power sector, they although observed bigger gain to opt for a robust strategy rather than the deterministic solution.

@@

Then, we have tested three roadmaps resulting from different perfect foresight optimisations: REF, SMR and ROB. Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}. From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation of the additional capacities needed in myopic optimisations subject to uncertainties. This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation. However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario. Even though this roadmap represents 7.4\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).  On top of decreasing the risk to invest into additional capacities, the ROB is also less sensitive to the variation of the costs of purchasing the energy carriers that are the main contributors of uncertainty.

### 2024-06-05 chap_atom_mol/chap_atom_mol.tex
Typos + Init Rebuttals

Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol are selected to supply, on average, 5\% of the demand in methanol. Its \gls{NED} represents, on average, 3\% of the total consumption of methnaol. The variation of imported e-methanol is due to its role in the industrial production of \gls{HVC} through the \acrfull{MTO} process as it represents 95\% of the total consumption. For the remaining 2\%, methanol is also used to supply the freight transport sector via boats or trucks. More detailed statistics including the different sources of supply and consumption of gas, hydrogen, ammonia and methanol are provided in Appendix \ref{app:UQ_electrofuels}.\\

This part assesses the space of uncertainties, like \citet{pickering2022diversity} who investigated the space of feasibility to reach carbon neutrality in Europe. The trend lines of the key parameters can be drawn for these imports and the installed capacities of \gls{SMR} in 2050. We picked 2050 as it is the time of the transition where electrofuels, if imported, are imported in the largest amount compared to the other years of the transiton.Next to the name of a parameter, one can read its Sobol' index versus the output of interest. For these outputs of interest, different from the total transition cost, the \gls{LOO} error is around 20\%. Even though this is higher than the threshold of \SI{1}{\%}, the Sobol' indices allow ranking the most impacting parameters (see Section \ref{subsec:pce}). Box plots also point out the distribution of this output for the extreme low or high values of some parameters.

>>>>

Likewise, e-methanol early becomes the selected option to supply methanol even though alternatives like biomass-to-methanol are selected to supply, on average, 5\% of the demand in methanol. Its \gls{NED} represents, on average, 3\% of the total consumption of methanol. The variation of imported e-methanol is due to its role in the industrial production of \gls{HVC} through the \acrfull{MTO} process as it represents 95\% of the total consumption. For the remaining 2\%, methanol is also used to supply the freight transport sector via boats or trucks. More detailed statistics including the different sources of supply and consumption of gas, hydrogen, ammonia and methanol are provided in Appendix \ref{app:UQ_electrofuels}.\\

This part assesses the space of uncertainties, like \citet{pickering2022diversity} who investigated the space of feasibility to reach carbon neutrality in Europe. The trend lines of the key parameters can be drawn for these imports and the installed capacities of \gls{SMR} in 2050. We picked 2050 as it is the time of the transition where electrofuels, if imported, are imported in the largest amount compared to the other years of the transition. Next to the name of a parameter, one can read its Sobol' index versus the output of interest. For these outputs of interest, different from the total transition cost, the \gls{LOO} error is around 20\%. Even though this is higher than the threshold of \SI{1}{\%}, the Sobol' indices allow ranking the most impacting parameters (see Section \ref{subsec:pce}). Box plots also point out the distribution of this output for the extreme low or high values of some parameters.

### 2024-06-05 chap_case_study/chap_case_study.tex
Typos + Init Rebuttals

In the current work, imported electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``renewable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}. In practice, it means that their \gls{GWP} is assumed to be zero in the model. Compared to equivalent electrofuels that would be locally produced, this assumption does not account account for the emissions related to their transport of imported electrofuels from abroad to Belgium. However, the share of these emissions is limited (between 5 and 10\%) compared to the whole cycle of production and use of imported electrofuels \cite{coppitters2024towards}. Therefore, the conclusions drawn in this thesis are not affected by this assumption.

@@

This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors. A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}. As detailed before, the commercial availability of such a technology is uncertain but would not be before 2040. Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology. Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% of chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward. }. Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about its availability. Finally, the the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.

>>>>

In the current work, imported electrofuels (\ie e-methane, e-hydrogen, e-methanol and e-ammonia) are assumed to be ``renewable" in the sense that they do not increase the concentration of \ce{CO2} in the atmosphere \cite{rixhon2021terminology}. In practice, it means that their \gls{GWP} is assumed to be zero in the model. Compared to equivalent electrofuels that would be locally produced, this assumption does not account for the emissions related to the transport of imported electrofuels from abroad to Belgium. However, the share of these emissions is limited (between 5 and 10\%) compared to the whole cycle of production and use of imported electrofuels \cite{coppitters2024towards}. Therefore, the conclusions drawn in this thesis are not affected by this assumption.

@@

This work considers nine groups of uncertain parameters: (i) the cost of purchasing imported energy carriers; (ii) the investment cost (\ie CAPEX) of some technologies, mostly related to the mobility sector and the integration of renewables; (iii) the maintenance cost (\ie OPEX) of every technology; (iv) the consumption of electric and fuel cells vehicles in the mobility sector; (v) the potential installed capacity of renewables; (vi) the hourly load factor of renewables accounting for variability of solar irradiance or wind speed; (vii) the availability of resources considered as limited (\ie biomass and electricity); (viii) the end-use-demands split per sector of activities (\ie households, services, passenger mobility and industry) and (ix) other parameters like the interest rate or the modal share change in different key sectors. A particular attention is to pay to the potential installation of \gls{SMR}, at the bottom of \Cref{tab:UC_full}. As detailed before, the commercial availability of such a technology is uncertain but would not be before 2040. Consequently, for \gls{SMR}, the parameter $f_{\mathrm{max,SMR}}$ influences the maximum capacity to install to translate somehow the readiness of this technology. Arbitrarily, we have then assumed the following probability of availability of such a technology: 10\% of chance to be installable from 2040, 20\% from 2045 and 40\% from 2050\footnote{In other words, if this parameter, ranging between 0 and 1, is (i) smaller than 0.6, there is no possibility to install \gls{SMR} during the transition; (ii) between 0.6 and 0.8, these 6~GW can be installed only in 2050; (iii) between 0.8 and 0.9, these can be installed from 2045 onward and; (iv) higher than 0.9, the prescribed maximum capacity can be installed from 2040 onward. }. Based on the local sensitivity analysis carried out by \citet{PATHS2050}, the current work also considers a [-40\%; +44\%] range on the CAPEX of SMR, on top of the uncertainty about its availability. Finally, the cost of purchasing renewable electrofuels presents a wide range, [-64.3\%; +179.8\%], like the other imported commodities.

### 2024-06-05 chap_intro_ccl/conclusion.tex
Typos + Init Rebuttals

About the use of the electrofuels, there are two molecules imported in larger quantities: e-methanol and e-ammonia. E-ammonia would aim at defossilising the production of \acrfull{HVC} in the non-energy sector and e-ammonia would dominantly provide flexibility in the production of electricity, respectively. The main use of e-hydrogen is in the freight transport via trucks and e-methane would supply the industrial sectors through \gls{CHP} and boilers.

>>>>

About the use of the electrofuels, there are two molecules imported in larger quantities: e-methanol and e-ammonia. E-methanol would aim at defossilising the production of \acrfull{HVC} in the non-energy sector and e-ammonia would dominantly provide flexibility in the production of electricity, respectively. The main use of e-hydrogen is in the freight transport via trucks and e-methane would supply the industrial sectors through \gls{CHP} and boilers.
